---
title: "2.1 Preprocess metadata"
format:
  html:
    df-print: paged
    embed-resources: true
    toc: true
  gfm: default
---

## Purpose

This script builds `metadata.tsv`, the public metadata file every other script reads. It takes the raw CoRIS extract, the Olink/RNA-seq ID mappings, and the case-control pairing file, assigns pseudonymous IDs (sid, sid_pair), cleans some variables, and computes the time-to-event variables. Variables with re-identification risk (age, HIV transmission route, country of origin, ART start period) are kept as empty placeholder columns, so downstream scripts still run without those values being exposed.

## Reproducibility

This script needs five private, controlled-access source files and can't be run from the public repository. It's included so the derivation of `metadata.tsv` is auditable, not so it can be re-executed.

```{r}
#| message: false
library(haven)
library(lubridate)
library(readxl)
library(tidyverse)
```

## Source files

```{r}
#| message: false
id_group_pbmc <- read_xlsx("../restricted_data/id_group_PBMC.xlsx")
id_matching <- read_xlsx("../restricted_data/id_matching_controls_cases.xlsx") # columns: ID_event, ID_control
PRETTY_metadata <- read_dta("../restricted_data/PRETTY_metadata.dta") # 187 samples, 68 variables
```

## Fix database issues

- `death_all`: `NA` means alive/no event (`death_all == 0`). Additionally correct a database inconsistency for participant `id == 9935`, who is recorded as not deceased (`death_all == 0`) being a control, when all the controls are recorded as `NA`.

```{r}
# Use labels from Stata file in the group category
PRETTY_metadata$group <- haven::as_factor(PRETTY_metadata$group)
PRETTY_metadata |> count(group, death_all)
```

```{r}
PRETTY_metadata |> 
    filter(group == "Control", death_all == 0) |> 
    select(id, group, death_all)
```

```{r}
PRETTY_metadata <- PRETTY_metadata |>
    mutate(death_all = if_else(is.na(death_all) | id == 9935, 
                               0, death_all))
# Remove Stata labels
PRETTY_metadata$death_all <- as.numeric(PRETTY_metadata$death_all)
```

```{r}
PRETTY_metadata |> count(group, death_all)
```

## Add Olink and RNA-seq sample IDs

```{r}
PRETTY_metadata$olink_id <- PRETTY_metadata$id_muestra
```

```{r}
id_group_pbmc <- id_group_pbmc |>
    mutate(rnaseq_id = gsub("-", "_", IDTUBO)) |>
    select(id_sample, rnaseq_id)

PRETTY_metadata <- PRETTY_metadata |>
    left_join(id_group_pbmc, by = c("id_muestra" = "id_sample"))
```

## Pseudonymize: assign `sid`

```{r}
PRETTY_metadata <- PRETTY_metadata |>
    mutate(sid = paste0("PTY_", sprintf("%03d", row_number())))
```

## Matched-pair identifier (`sid_pair`)

```{r}
id_matching_long <- bind_rows(
    id_matching |> mutate(id = ID_event, id_pair = ID_control,
                          .keep = "none"),
    id_matching |> mutate(id = ID_control, id_pair = ID_event,
                             .keep = "none"),
)

PRETTY_metadata <- PRETTY_metadata |>
    left_join(id_matching_long, by = "id") |>
    mutate(sid_pair = sid[match(id_pair, id)]) |>
    select(-id_pair)
```

## Time to event (`time`)

```{r}
PRETTY_metadata <- PRETTY_metadata |>
    mutate(
        time = if_else(
            group == "Event",
            floor(interval(f_muestra24, fechaENOS) / months(1)),
            floor(interval(f_muestra24, art_endfu) / months(1))
        )
    )
```

## Months from ART initiation to sampling (`months_art_sampling`)

```{r}
PRETTY_metadata <- PRETTY_metadata |>
    mutate(
        months_art_sampling = time_length(interval(art_ini, dmy(Fechadeobtención)), unit = "month")
    )
```

## De-identified placeholder columns

`age`, `mode` (HIV acquisition risk factor), `paisorigen` (country of origin), and `artini_period` (year of ART initiation) carry elevated re-identification risk. Therefore, they are included as `NA` columns so other scripts can still run their full table structure publicly.

```{r}
PRETTY_metadata <- PRETTY_metadata |>
    mutate(age = NA, mode = NA, paisorigen = NA, artini_period = NA)
```

## Recode `group`, `sex`, etc. to the labels used publicly

```{r}
PRETTY_metadata <- PRETTY_metadata |>
    haven::as_factor() |>
    mutate(
        sex = fct_recode(sex, "Male" = "Hombre", "Female" = "Mujer"),
        edulvl = fct_recode(
            edulvl,
            "No studies" = "Sin estudios o primaria incompleta",
            "Primary school" = "Primaria actual completa (hasta 5º EGB)",
            "Secondary school" = "Secundaria obligatoria completa (EGB/ESO)",
            "High school" = "Bachillerato completo (BUP/COU/FP)",
            "University" = "Universitarios o superiores",
            "Other" = "5"
        ),
        art_reg = fct_recode(
            art_reg,
            "2 NRTI + 1 NNRTI" = "2NRTI+1NNRTI",
            "2 NRTI + 1 PI" = "2NRTI+1IP",
            "2 NRTI + 1 INSTI" = "2NRTI+1II",
            "Other" = "otro"
        )
    )
```

## Final public metadata file

```{r}
metadata_public <- PRETTY_metadata |>
    select(
        sid, group, nadm, mace, death_all,
        age, sex, mode, paisorigen, edulvl,
        smoke, hta,
        art_reg,
        cvmax, nadircd4,
        CD4_0, CD8_0, CD4CD8_0,
        CD4_24m, CD8_24m, CD4CD8_24,
        months_art_sampling, artini_period,
        sid_pair, time
    )
```

```{r}
# metadata_public |> write_tsv("../data/metadata.tsv")
```

## Session info

```{r}
sessioninfo::session_info()
```
