---
title: "2.2 Preprocess and QC Olink"
format:
  html:
    df-print: paged
    embed-resources: true
    toc: true
  gfm: default
---

## Purpose

This script runs the Olink QC pipeline on the raw NPX data. It removes samples flagged with QC warnings, outliers, and unmatched pairs (using `sid_pair` to drop a flagged sample's partner too), then filters out assays with high missingness. It outputs `olink_npx_qced_sid.tsv`, the 138-sample, 326-protein dataset used by every downstream proteomics script, and appends the `olink_included`/`rnaseq_included` flags back into `metadata.tsv`.

## Reproducibility

This script runs entirely from the public repository. Only `olink_proteomics_npx.csv` and `metadata.tsv` are needed.

```{r}
#| message: false
library(OlinkAnalyze)
library(tidyverse)
```

In the Olink report, we have the following QC relevant information, that we will take as starting point:

-   Olink Panel: Explore 384 Inflammation
-   Normalization method: Intensity normalization (except PNLIPRP2)
-   Number of samples: 187
-   Samples passed QC: 171
-   Number of proteins: 368
-   Number of detected proteins: 338

## Read NPX data

This file is the same raw Olink export deposited in PRIDE. Sample identifiers here are already the pseudonymous `sid` (`PTY_XXX`), except for Olink's own kit control samples.

```{r}
pdata <- read_NPX("../data/olink_proteomics_npx.csv")
```

Remove Olink's internal kit control samples (not participant data):

```{r}
pdata <- pdata |> filter(!grepl("CONTROL", SampleID))
```

`SampleID` already contains `sid` for every remaining row. We keep the `SampleID` column intact (OlinkAnalyze functions like `olink_qc_plot()` and `olink_ttest()` require a column with this exact name internally) and add `sid` as an alias for our own filtering/joining below.

```{r}
pdata <- pdata |> mutate(sid = SampleID)
```

Add information about the groups:

```{r}
#| message: false
mdata_groups <- read_tsv("../data/metadata.tsv") |>
    select(sid, group, nadm, mace, death_all)
pdata <- pdata |> left_join(mdata_groups, by = "sid")
```

## Number of measurements

First, we count the number of measurements in each group:

```{r}
pdata |> count(group) |> mutate(pts = n / 368)
```

We have 95 patients in the control group and 92 patients in the events group. There are 3 patients that do not have a correct PS-matched pair in the control group. We will remove these samples from further analyses.

To identify a flagged sample's matched partner during QC, we need the case-control pairing structure. This is provided as the `sid_pair` column in `metadata.tsv`.

```{r}
#| message: false
id_matching_pairs <- read_tsv("../data/metadata.tsv") |>
    select(sid, sid_pair)
```

## Samples quality control

### Filter 1: QC Warning

Remove samples with Warning QC status:

```{r}
sample_warn <- pdata |>
    filter(QC_Warning == "WARN") |>
    count(sid) |>
    mutate(percentage = round(n/n_distinct(pdata$Assay)*100)) |>
    arrange(desc(percentage))
sample_warn_ids <- sample_warn |> pull(sid)
sample_warn
```

Check pairs of these samples:

```{r}
sample_warn_ids_pairs <- id_matching_pairs |> filter(sid %in% sample_warn_ids) |> pull(sid_pair)
sample_warn_remove_all <- unique(c(sample_warn_ids, sample_warn_ids_pairs))
```

```{r}
# Remove samples
pdata_filter1 <- pdata |> filter(!sid %in% sample_warn_remove_all)
```

This is the total number of remaining samples:

```{r}
pdata_filter1 |> count(group) |> mutate(pts = n / 368)
```

### Filter 2: Outliers

Identify outliers:

```{r}
qc <- olink_qc_plot(pdata_filter1)
outliers <- qc$data |> filter(Outlier == 1) |> pull(SampleID)
outliers_sid <- pdata |> select(SampleID, sid) |> distinct() |> filter(SampleID %in% outliers) |> pull(sid)
```

5 outliers were identified. We are going to remove these 5 and their paired participants.

Identify pairs of these samples:

```{r}
outliers_ids_pairs <- id_matching_pairs |> filter(sid %in% outliers_sid) |> pull(sid_pair)
outliers_remove_all <- unique(c(outliers_sid, outliers_ids_pairs))
```

Remove:

```{r}
pdata_filter2 <- pdata_filter1 |> filter(!sid %in% outliers_remove_all)
```

This is the total number of remaining samples:

```{r}
pdata_filter2 |> count(group) |> mutate(pts = n / 368)
```

### Filter 3: Unmatched samples

Identify unmatched samples (no valid PS-matched pair):

```{r}
unmatched_sid <- id_matching_pairs |> filter(is.na(sid_pair)) |> pull(sid)
```

Remove:

```{r}
pdata_filter3 <- pdata_filter2 |> filter(!sid %in% unmatched_sid)
```

This is the total number of remaining samples:

```{r}
pdata_filter3 |> count(group) |> mutate(pts = n / 368)
```

## Assays quality control

### Filter 4: Missingness

Identify assays with high missingness:

```{r}
assays_todel <- pdata_filter3 |>
    mutate(Missing = ifelse(NPX > LOD, FALSE, TRUE)) |>
    group_by(Assay, group) |>
    summarise(Missing = round(mean(Missing), 4 ) > 0.25) |>
    group_by(Assay) |>
    summarise(Missing_status = all(Missing)) |>
    filter(Missing_status == TRUE) |>
    pull(Assay)
assays_todel
```

Remove:

```{r}
pdata_filter4 <- pdata_filter3 |> filter(!Assay %in% assays_todel)
```

### Filter 5: Assay Warning

Count assay warnings:

```{r}
pdata_filter4 |> count(Assay_Warning)
```

There are no flagged assays.

## Final dataset

```{r}
pdata_final <- pdata_filter4
```

The final dataset contains 138 samples and 326 proteins, distributed in the following groups:

```{r}
pdata_final |> select(sid, group) |> distinct() |> count(group)
```

```{r}
pdata_final |> select(sid, nadm) |> distinct() |> count(nadm)
```

```{r}
pdata_final |> select(sid, mace) |> distinct() |> count(mace)
```

```{r}
pdata_final |> select(sid, death_all) |> distinct() |> count(death_all)
```

Save to file:

```{r}
# pdata_final |> write_tsv("../data/olink_npx_qced_sid.tsv")
```

Add qc columns to `metadata.tsv`:

```{r}
#| message: false
olink_qced_sids <- pdata_final |> pull(sid) |> unique()
rnaseq_sids <- colnames(read_tsv("../data/rnaseq_transcriptomics_counts.tsv"))[-1]

metadata <- read_tsv("../data/metadata.tsv") |>
    mutate(
        olink_included = as.integer(sid %in% olink_qced_sids),
        rnaseq_included = as.integer(sid %in% rnaseq_sids)
    )

# metadata |> write_tsv("../data/metadata.tsv")
```

## Session info

```{r}
sessioninfo::session_info()
```
