---
title: "4. Differential expression analysis of proteins"
format:
  html:
    df-print: paged
    embed-resources: true
    toc: true
  gfm: default
---

## Purpose

This script runs differential expression analysis on the Olink proteomics data across the four outcomes (SNAE, malignancy, cardiovascular event, mortality) using `olink_ttest`. It produces the `dep_*.csv` files used by the mixomics and correlation scripts, and generates the FDR-adjusted volcano plots (Fig. 1).

## Reproducibility

This script runs entirely from the public repository. Only `olink_npx_qced_sid.tsv` is needed.

```{r}
#| message: false
library(ggpubr)
library(ggrepel)
library(OlinkAnalyze)
library(tidyverse)
```

## Read data

```{r}
#| message: false
pdata <- read_tsv("../data/olink_npx_qced_sid.tsv") |>
    mutate(
        group = factor(group, levels = c("Event", "Control")),
        nadm = factor(nadm, levels = c(1, 0)),
        mace = factor(mace, levels = c(1, 0)),
        death_all = factor(death_all, levels = c(1, 0))
    )
```

## Shared helpers

Colours/alphas and the differential-expression classification are identical across every outcome. The FDR volcano plot is also shared, but each outcome has small differences from the others (whether the FDR=0.05 reference line is drawn, y-axis expansion, label nudge/force, and whether one assay's label needs a separately-nudged position to avoid overlap). Those are exposed as parameters rather than hardcoded, so each call site states exactly how it differs.

```{r}
cols   <- c("Up" = "#ffad73", "Down" = "#26b3ff", "n.s." = "grey")
alphas <- c("Up" = 1, "Down" = 1, "n.s." = 0.5)
highlight_assays <- c("LGALS9", "DAPP1", "SKAP2", "CCL13")

compute_dep <- function(pdata, outcome_var) {
    olink_ttest(pdata, outcome_var) |>
        mutate(prot_exp = case_when(
            estimate >= 0 & Adjusted_pval <= 0.05 ~ "Up",
            estimate <= 0 & Adjusted_pval <= 0.05 ~ "Down",
            TRUE ~ "n.s."
        ))
}

plot_volcano_fdr <- function(dep_df, title, xlim_range,
                             show_hline = TRUE,
                             y_expand = NULL,
                             label_nudge_y = 0.05,
                             label_force = 15,
                             split_label = NULL) {
    p <- dep_df |>
        ggplot(aes(x = estimate, y = -log10(Adjusted_pval), fill = prot_exp, alpha = prot_exp)) +
        geom_point(shape = 21, colour = "black")
    
    if (show_hline) {
        p <- p + geom_hline(yintercept = -log10(0.05), linetype = "dashed")
    }
    if (!is.null(y_expand)) {
        p <- p + scale_y_continuous(expand = expansion(mult = y_expand))
    }
    
    # Main label layer -- excludes split_label$assay when one assay needs a
    # separately-nudged position (e.g. SKAP2 in the NADM plot).
    label_data <- if (!is.null(split_label)) dep_df |> filter(Assay != split_label$assay) else dep_df
    p <- p +
        geom_label_repel(
            data = label_data,
            aes(label = if_else(Assay %in% highlight_assays, Assay, "")),
            size = 2, show.legend = FALSE, max.overlaps = Inf,
            nudge_y = label_nudge_y, min.segment.length = 0,
            force = label_force, label.padding = unit(0.1, "lines")
        )
    
    if (!is.null(split_label)) {
        p <- p +
            geom_label_repel(
                data = dep_df |> filter(Assay == split_label$assay),
                aes(label = if_else(Assay %in% highlight_assays, Assay, "")),
                size = 2, show.legend = FALSE, max.overlaps = Inf,
                nudge_y = split_label$nudge_y, nudge_x = split_label$nudge_x,
                min.segment.length = 0, force = 15, label.padding = unit(0.1, "lines")
            )
    }
    
    p +
        scale_fill_manual(values = cols, breaks = c("Up", "Down", "n.s."),
                          labels = c("Up", "Down", "n.s.")) +
        scale_alpha_manual(values = alphas) +
        xlim(xlim_range) +
        theme_bw() +
        theme(legend.position = "bottom", legend.background = element_rect(colour = "black"),
              axis.title = element_text(size = 9)) +
        labs(fill = "Differential abundance", title = title,
             x = expression(log[2](FC)), y = expression(-log[10]("FDR-adjusted P value"))) +
        guides(alpha = "none", fill = guide_legend(override.aes = list(alpha = 1)))
}
```

## Group

```{r}
dep_group <- compute_dep(pdata, "group")
dep_group
# dep_group |> write_tsv("../results/tableS2_dep_group.tsv")
```

```{r}
p1 <- plot_volcano_fdr(dep_group, "PWH with vs without SNAEs", xlim_range = c(-1.1, 1.1),
                       y_expand = c(0.02, 0.15), label_nudge_y = 0.1, label_force = 1)
p1
```

## NADM

```{r}
dep_nadm <- compute_dep(pdata, "nadm")
# dep_nadm |> write_tsv("../results/tableS3_dep_nadm.tsv")
```

```{r}
p2 <- plot_volcano_fdr(dep_nadm, "PWH with vs without neoplasias", xlim_range = c(-1.7, 1.7),
                       split_label = list(assay = "SKAP2", nudge_y = 0.08, nudge_x = -0.3))
p2
```

## MACE

```{r}
dep_mace <- compute_dep(pdata, "mace")
# dep_mace |> write_tsv("../results/tableS4_dep_mace.tsv")
```

```{r}
p3 <- plot_volcano_fdr(dep_mace, "PWH with vs without CV events", xlim_range = c(-1.2, 1.2),
                       show_hline = FALSE)
p3
```

## Death_all

```{r}
dep_death_all <- compute_dep(pdata, "death_all")
# dep_death_all |> write_tsv("../results/tableS5_dep_death_all.tsv")
```

```{r}
p4 <- plot_volcano_fdr(dep_death_all, "PWH with vs without mortality", xlim_range = c(-2.8, 2.8))
p4
```

```{r}
# svg("../results/fig1_AtoD.svg", width = 7, height = 5)
ggarrange(p1, p2, p3, p4, labels = c("A", "B", "C", "D"),
          common.legend = TRUE, legend = "bottom")
# dev.off()
```

## Session info

```{r}
sessioninfo::session_info()
```
