---
title: "5. RNA-seq analysis"
format:
  html:
    df-print: paged
    embed-resources: true
    toc: true
  gfm: default
---

## Purpose

This script runs differential expression analysis on the RNA-seq data across the four outcomes using DESeq2, producing the `deg_*.tsv` files (Table S7 to S10) and the FDR-adjusted volcano plots (Fig. 2). It also runs the cell-type deconvolution statistical analysis (Wilcoxon tests per cell type, BH-corrected within each outcome), reproducing Table S11 and Fig. S2.

## Reproducibility

This script runs entirely from the public repository. Only `rnaseq_transcriptomics_counts.tsv`, `metadata.tsv`, and `cell_deconvolution_sid.csv` are needed.

```{r}
#| message: false
library(AnnotationDbi)
library(biomaRt)
library(org.Hs.eg.db)
library(DESeq2)
library(ggpubr)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(tidyverse)
```

## Data generation

RNA-seq reads were processed with nf-core/rnaseq v3.17.0, using STAR/RSEM for alignment and quantification, against the GRCh38.112 Ensembl reference. `rnaseq_transcriptomics_counts.tsv` is derived from the pipeline's `rsem.merged.gene_counts.tsv` output.

```{bash}
#| eval: false
nextflow run \
    nf-core/rnaseq -r 3.17.0 \
    --input "samplesheet.csv" \
    --outdir "results" \
    --gtf "Homo_sapiens.GRCh38.112.gtf.gz" \
    --fasta "Homo_sapiens.GRCh38.dna_sm.primary_assembly.fa.gz" \
    --aligner "star_rsem" \
    --igenomes_ignore \
    --genome null \
    -profile singularity \
    -c nextflow.config
```

Reference files (`Homo_sapiens.GRCh38.112.gtf.gz`, `Homo_sapiens.GRCh38.dna_sm.primary_assembly.fa.gz`) are the standard Ensembl release 112 GRCh38 assembly, publicly available from Ensembl.

## Load public data and build the dds object

Raw RNA-seq counts (genes x samples) and the public metadata file are combined here to build the `DESeqDataSet` from scratch.

```{r}
#| message: false
counts_mat <- read_tsv("../data/rnaseq_transcriptomics_counts.tsv") |>
    column_to_rownames("gene_ensembl_id") |>
    as.matrix()
```

```{r}
#| message: false
coldata <- read_tsv("../data/metadata.tsv") |>
    filter(sid %in% colnames(counts_mat)) |>
    column_to_rownames("sid")

coldata <- coldata[colnames(counts_mat), , drop = FALSE] # match sample order
```

```{r}
stopifnot(all(colnames(counts_mat) == rownames(coldata))) # Must be TRUE
```

```{r}
dds <- DESeqDataSetFromMatrix(
    countData = round(counts_mat),
    colData   = coldata,
    design    = ~1
)
```

## dds inspection

```{r}
head(counts(dds)) # to check the raw counts
```

```{r}
colData(dds) # to check the sample info
```

```{r}
design(dds) # to check the design formula
```

## Pre-filtering

```{r}
smallestGroupSize <- 3

keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds_filtered <- dds[keep, ]
```

## Help functions

```{r}
# A function to run DESeq2
run_deseq <- function(dds, design_var) {
    dds_tmp <- DESeqDataSet(dds, design = as.formula(paste0("~ ", design_var)))
    dds_tmp <- DESeq(dds_tmp)
    res <- results(dds_tmp)
    list(dds = dds_tmp, res = res)
}
```

```{r}
# A function for Ensembl annotation
annotate_genes <- function(ensembl_ids) {
    mart <- useEnsembl("genes", "hsapiens_gene_ensembl")
    
    getBM(
        attributes = c(
            "ensembl_gene_id",
            "external_gene_name",
            "description",
            "gene_biotype"
        ),
        filters = "ensembl_gene_id",
        values = unique(ensembl_ids),
        mart = mart
    )
}
```

```{r}
tidy_results <- function(res, lfc = 0, alpha = 0.05, filter_sig = TRUE) {
    
    # 1. Convert DESeq2 results to tibble and classify genes
    res_tb <- as_tibble(res) |>
        mutate(
            gene = rownames(res),
            diffexpressed = case_when(
                log2FoldChange >  lfc & padj < alpha ~ "upregulated",
                log2FoldChange < -lfc & padj < alpha ~ "downregulated",
                TRUE ~ "not_de"
            )
        ) |>
        relocate(gene)
    
    # 2. Annotate ALL genes
    annot <- annotate_genes(res_tb$gene) |>
        mutate(
            description = sub("\\s*\\[.*\\]$", "", description)
        )
    
    # 3. Join annotation
    res_tb <- res_tb |>
        left_join(
            annot,
            by = c("gene" = "ensembl_gene_id")
        ) |>
        arrange(pvalue)
    
    # 4. Optionally filter significant genes
    if (filter_sig) {
        res_tb <- res_tb |> filter(padj < alpha)
    }
    
    return(res_tb)
}
```

```{r}
# FDR-adjusted volcano plot
plot_volcano_fdr <- function(
        res_tb,
        title,
        lfc_cutoff = 1,
        fdr_cutoff = 0.05,
        label_pc_only = TRUE
) {
    cols <- c("upregulated" = "#ffad73", "downregulated" = "#26b3ff", "not_de" = "grey")
    alphas <- c("upregulated" = 1, "downregulated" = 1, "not_de" = 0.5)
    
    plot_tb <- res_tb |>
        mutate(
            genelabels = ifelse(
                diffexpressed != "not_de" &
                    (!label_pc_only | gene_biotype == "protein_coding"),
                external_gene_name,
                ""
            )
        )
    xmax <- max(abs(plot_tb$log2FoldChange), na.rm = TRUE)
    
    ggplot(
        plot_tb,
        aes(x = log2FoldChange, y = -log10(padj), fill = diffexpressed, alpha = diffexpressed)
    ) +
        geom_point(shape = 21, colour = "black") +
        geom_vline(xintercept = c(-lfc_cutoff, lfc_cutoff), linetype = "dashed",
                   linewidth = 0.2, colour = "black") +
        geom_hline(yintercept = -log10(fdr_cutoff), linetype = "dashed") +
        scale_fill_manual(values = cols, name = "Differential expression",
                          breaks = c("upregulated", "downregulated", "not_de"),
                          labels = c("Up", "Down", "n.s.")) +
        scale_alpha_manual(values = alphas) +
        scale_x_continuous(limits = c(-xmax, xmax)) +
        theme_bw() +
        theme(legend.position = "bottom",
              legend.background = element_rect(colour = "black", linewidth = 0.5),
              axis.title = element_text(size = 9)) +
        labs(title = title, x = expression(log[2](FC)),
             y = expression(-log[10]("FDR-adjusted P value"))) +
        guides(alpha = "none")
}
```

## Run analyses

```{r}
res_group      <- run_deseq(dds_filtered, "group")
res_nadm       <- run_deseq(dds_filtered, "nadm")
res_mace       <- run_deseq(dds_filtered, "mace")
res_death_all  <- run_deseq(dds_filtered, "death_all")
```

## Tidy + plot

```{r}
tb_group <- tidy_results(res_group$res, filter_sig = FALSE)
tb_group
tb_nadm  <- tidy_results(res_nadm$res, filter_sig = FALSE)
tb_nadm
tb_mace  <- tidy_results(res_mace$res, filter_sig = FALSE)
tb_mace
tb_death <- tidy_results(res_death_all$res, filter_sig = FALSE)
tb_death
```

## Export results

```{r}
# tb_group |> write_tsv("../results/tableS7_deg_group.tsv")
# tb_nadm |> write_tsv("../results/tableS8_deg_nadm.tsv")
# tb_mace |> write_tsv("../results/tableS9_deg_mace.tsv")
# tb_death |> write_tsv("../results/tableS10_deg_death_all.tsv")
```

## Volcano plots

```{r}
pA <- plot_volcano_fdr(tb_group, "PWH with vs without SNAEs")
pA
pB <- plot_volcano_fdr(tb_nadm, "PWH with vs without neoplasias")
pB
pC <- plot_volcano_fdr(tb_mace, "PWH with vs without CV events")
pC
pD <- plot_volcano_fdr(tb_death, "PWH with vs without mortality")
pD
```

All together in one figure:

```{r}
# svg("../results/fig2_AtoD.svg", width = 7, height = 5)
ggarrange(pA, pB, pC, pD, labels = c("A", "B", "C", "D"),
          common.legend = TRUE, legend = "bottom")
# dev.off()
```

## Cell deconvolution statistical analysis

This section reproduces the statistics reported in Table S11/Fig. S2 directly from the public deconvolution results file.

```{r}
#| message: false
deconv <- read_csv("../data/cell_deconvolution_sid.csv")
metadata <- read_tsv("../data/metadata.tsv")

outcomes <- c("group", "nadm", "mace", "death_all")
diagnostic_columns <- c("sid", "P-value", "Correlation", "RMSE")
cell_types <- setdiff(names(deconv), diagnostic_columns)
```

```{r}
fit_qc <- deconv |>
    summarise(
        samples = n(),
        cibersort_p_lt_0.001 = sum(`P-value` < 0.001),
        correlation_median = median(Correlation),
        correlation_q1 = quantile(Correlation, 0.25),
        correlation_q3 = quantile(Correlation, 0.75),
        rmse_median = median(RMSE),
        rmse_q1 = quantile(RMSE, 0.25),
        rmse_q3 = quantile(RMSE, 0.75)
    )
fit_qc
```

```{r}
celltype_qc <- deconv |>
    dplyr::select(all_of(cell_types)) |>
    pivot_longer(everything(), names_to = "cell_type", values_to = "cell_proportion") |>
    group_by(cell_type) |>
    summarise(
        mean_proportion = mean(cell_proportion),
        median_proportion = median(cell_proportion),
        q1_proportion = quantile(cell_proportion, 0.25),
        q3_proportion = quantile(cell_proportion, 0.75),
        nonzero_n = sum(cell_proportion > 0),
        zero_n = sum(cell_proportion == 0),
        zero_percent = 100 * mean(cell_proportion == 0),
        .groups = "drop"
    ) |>
    arrange(desc(mean_proportion))
celltype_qc
```

```{r}
# group/nadm/mace/death_all: convert group's "Control"/"Event" labels to
# 0/1 first, so all four outcomes share the same 0/1 convention.
metadata_rnaseq <- metadata |>
    filter(sid %in% deconv$sid) |>
    mutate(group = if_else(group == "Event", 1, 0)) |>
    mutate(across(all_of(outcomes), as.integer))

analysis_data <- deconv |>
    left_join(metadata_rnaseq, by = "sid")
```

```{r}
sample_counts <- analysis_data |>
    dplyr::select(sid, all_of(outcomes)) |>
    pivot_longer(cols = all_of(outcomes), names_to = "outcome", values_to = "event") |>
    filter(!is.na(event)) |>
    dplyr::count(outcome, event, name = "n")
sample_counts
```

```{r}
deconv_long <- analysis_data |>
    dplyr::select(sid, all_of(cell_types), all_of(outcomes)) |>
    pivot_longer(cols = all_of(cell_types), names_to = "cell_type", values_to = "cell_proportion") |>
    pivot_longer(cols = all_of(outcomes), names_to = "outcome", values_to = "event") |>
    filter(!is.na(event))

compare_groups <- function(data) {
    event0 <- data$cell_proportion[data$event == 0]
    event1 <- data$cell_proportion[data$event == 1]
    
    test <- suppressWarnings(wilcox.test(event1, event0, paired = FALSE, exact = FALSE, correct = TRUE))
    
    combined_ranks <- rank(c(event1, event0), ties.method = "average")
    n1 <- length(event1); n0 <- length(event0)
    mann_whitney_u <- sum(combined_ranks[seq_len(n1)]) - n1 * (n1 + 1) / 2
    rank_biserial <- (2 * mann_whitney_u / (n1 * n0)) - 1
    
    tibble(
        n_event0 = n0, n_event1 = n1,
        median_event0 = median(event0), q1_event0 = quantile(event0, 0.25, names = FALSE),
        q3_event0 = quantile(event0, 0.75, names = FALSE),
        median_event1 = median(event1), q1_event1 = quantile(event1, 0.25, names = FALSE),
        q3_event1 = quantile(event1, 0.75, names = FALSE),
        median_difference = median(event1) - median(event0),
        rank_biserial = rank_biserial, p_value = test$p.value
    )
}

celltype_results <- deconv_long |>
    group_by(outcome, cell_type) |>
    group_modify(~ compare_groups(.x)) |>
    ungroup() |>
    group_by(outcome) |>
    mutate(p_adj = p.adjust(p_value, method = "BH")) |>
    ungroup() |>
    arrange(outcome, p_adj, p_value)
celltype_results
```

```{r}
nominal_findings <- celltype_results |> filter(p_value < 0.05)
fdr_findings <- celltype_results |> filter(p_adj < 0.05)
nominal_findings
fdr_findings
```

```{r}
# celltype_results |> write_tsv("../results/tableS11_cell_deconvolution_results.tsv")
```

## Deconvolution figure (Fig. S2)

```{r}
group_plot_data <- deconv_long |>
    filter(outcome == "group") |>
    mutate(
        event = factor(event, levels = c(0, 1), labels = c("No event", "Event")),
        cell_type = as.character(cell_type)
    )

cell_order <- group_plot_data |>
    group_by(cell_type) |>
    summarise(mean_abundance = mean(cell_proportion), .groups = "drop") |>
    arrange(desc(mean_abundance)) |>
    pull(cell_type)

group_plot_data <- group_plot_data |>
    mutate(cell_type = factor(cell_type, levels = cell_order))

label_positions <- group_plot_data |>
    mutate(cell_type = as.character(cell_type)) |>
    group_by(cell_type) |>
    summarise(y_position = max(cell_proportion) * 1.12 + 0.002, .groups = "drop")

nominal_labels <- celltype_results |>
    filter(outcome == "group", p_value < 0.05) |>
    dplyr::select(cell_type, p_value) |>
    mutate(cell_type = as.character(cell_type)) |>
    left_join(label_positions, by = "cell_type") |>
    mutate(
        cell_type = factor(cell_type, levels = cell_order),
        p_label = format.pval(p_value, digits = 1, eps = 0.001)
    )

deconvolution_figure <- ggplot(group_plot_data, aes(x = cell_type, y = cell_proportion, fill = event)) +
    geom_boxplot(position = position_dodge2(width = 0.8, preserve = "single"),
                 width = 0.68, outlier.size = 0.7, outlier.alpha = 0.5, linewidth = 0.35) +
    geom_text(data = nominal_labels, aes(x = cell_type, y = y_position, label = p_label),
              inherit.aes = FALSE, size = 3.2, fontface = "bold") +
    scale_fill_manual(values = c("No event" = "#0072B2", "Event" = "#D55E00")) +
    scale_y_sqrt(labels = scales::label_percent(accuracy = 1),
                 expand = expansion(mult = c(0.02, 0.14))) +
    labs(x = NULL, y = "Estimated cell proportion", fill = NULL) +
    theme_bw(base_size = 10) +
    theme(
        legend.position = "top",
        legend.background = element_rect(colour = "black", fill = "white", linewidth = 0.4),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 55, hjust = 1, vjust = 1, size = 8),
        axis.title.y = element_text(margin = margin(r = 8)),
        plot.caption = element_text(hjust = 0, size = 8)
    )

deconvolution_figure

# svg("../results/figS2_cell_deconvolution.svg", width = 7, height = 7)
deconvolution_figure
# dev.off()
```

## Session info

```{r}
sessioninfo::session_info()
```
