---
title: "8. Multiomics integration (DIABLO)"
format:
  html:
    df-print: paged
    embed-resources: true
    toc: true
  gfm: default
---

## Purpose

This script runs the DIABLO multi-omic integration analysis on the proteomics and transcriptomics data. It produces the four panels of Fig. S3, the sample plot, the top loadings, the circos plot, and the correlation heatmap restricted to the selected features.

## Reproducibility

This script runs entirely from the public repository, using `olink_npx_qced_sid.tsv`, `rnaseq_transcriptomics_vst.tsv`, and `dep_group.tsv`. The component and variable selection tuning steps are slow and are included as `eval: false`, with the resulting values hardcoded in the final model. Gene symbol annotation queries Ensembl through biomaRt at render time, so an internet connection is required.

```{r}
#| message: false
library(biomaRt)
library(ComplexHeatmap)
library(circlize)
library(DESeq2)
library(ggpubr)
library(grid)
library(mixOmics)
library(tidyverse)
```

## Load data

Both public files are already keyed by `sid` directly, so no private ID crosswalk is needed.

```{r}
#| message: false
pdata <- read_tsv("../data/olink_npx_qced_sid.tsv")
mdata <- pdata |>
    dplyr::select(sid:death_all) |>
    distinct() |>
    arrange(sid)
```

```{r}
#| message: false
rna_mat <- read_tsv("../data/rnaseq_transcriptomics_vst.tsv") |>
    column_to_rownames("gene_ensembl_id") |>
    as.matrix()
```

```{r}
# Remove near-constant genes, keep top 5000 most variable
vars <- apply(rna_mat, 1, var)
rna_mat <- rna_mat[order(vars, decreasing = TRUE)[1:5000], ]
# DIABLO expects centered/scaled blocks, samples as rows
rna_mat <- scale(t(rna_mat))
```

## Create input for DIABLO

```{r}
pdata_mat <- pdata |>
    dplyr::select(sid, Assay, NPX) |>
    pivot_wider(names_from = Assay, values_from = NPX) |>
    arrange(sid)

full_samples <- intersect(pdata_mat$sid, rownames(rna_mat))

protein_mat <- as.matrix(pdata_mat[, -1])
rownames(protein_mat) <- pdata_mat$sid
protein_mat <- protein_mat[full_samples, ]
rna_mat <- rna_mat[full_samples, ]

X <- list(protein = scale(protein_mat), mRNA = rna_mat)

Y_group <- factor(mdata |> filter(sid %in% full_samples) |> pull(group), labels = c("No", "Yes"))
names(Y_group) <- mdata |> filter(sid %in% full_samples) |> pull(sid)
summary(Y_group)
```

## DIABLO

### Design matrix

```{r}
design <- matrix(c(0, 0.1,
                   0.1, 0), ncol = 2)
```

### Number of components

```{r}
#| eval: false
diablo_screen <- block.plsda(X, Y_group, ncomp = 5, design = design)
perf.diablo.pretty <- perf(diablo_screen,
                           validation = 'Mfold',
                           folds = 10,
                           nrepeat = 10,
                           BPPARAM = BiocParallel::MulticoreParam(workers = 5))
plot(perf.diablo.pretty)
perf.diablo.pretty$choice.ncomp$WeightedVote
```

```{r}
# Result of the tuning above
ncomp <- 2
```

### Number of variables to select

Slow (grid search with repeated cross-validation) -- not run by default. The manuscript's Methods state 25 variables per block on 2 components, matching the hardcoded values used below.

```{r}
#| eval: false
test.keepX <- list(mRNA = c(5:9, seq(10, 25, 5)),
                   protein = c(5:9, seq(10, 25, 5)))
tune.diablo.pretty <- tune.block.splsda(X, Y_group, ncomp = 2,
                                        test.keepX = test.keepX, design = design,
                                        validation = 'Mfold', folds = 10, nrepeat = 100,
                                        dist = "max.dist",
                                        BPPARAM = BiocParallel::MulticoreParam(workers = 8))
tune.diablo.pretty$choice.keepX
```

```{r}
# Result of the tuning above -- matches the Methods (25 variables/block, 2 components)
list.keepX <- list(protein = c(25, 25), mRNA = c(25, 25))
```

### Final model

```{r}
diablo.pretty <- block.splsda(X, Y_group, ncomp = ncomp,
                              keepX = list.keepX, design = design)
```

## Sample plot (Fig. S3A)

```{r}
plotDiablo(diablo.pretty, ncomp = 1)
```

```{r}
# svg("../results/figS3a_diablo_sample_plot.svg", width = 7, height = 7)
plotIndiv(diablo.pretty, ind.names = FALSE, legend = TRUE, ellipse = TRUE,
          title = '', legend.title = "SNAEs")
# dev.off()
```

## Top 25 contributing variables per block (Fig. S3B)

```{r}
comp1_vars <- selectVar(diablo.pretty, comp = 1)
comp1_vars
```

Gene-symbol labels (the mRNA block is Ensembl-ID-keyed by default, but the manuscript names contributors by symbol):

```{r}
#| message: false
mart <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl", version = 112)

gene_map <- getBM(
    attributes = c("ensembl_gene_id", "hgnc_symbol"),
    filters = "ensembl_gene_id",
    values = comp1_vars$mRNA$name,
    mart = mart
) |>
    distinct(ensembl_gene_id, .keep_all = TRUE)

# A few IDs have no current HGNC symbol; assigned manually for figure legibility
manual_symbols <- c(
    "ENSG00000254802" = "lnc-ASPH-5",
    "ENSG00000261790" = "lnc-NOXO1-1",
    "ENSG00000285535" = "lnc-unnamed"
)

gene_map <- gene_map |>
    mutate(hgnc_symbol = if_else(ensembl_gene_id %in% names(manual_symbols),
                                 unname(manual_symbols[ensembl_gene_id]), hgnc_symbol))

mrna_ids <- colnames(X$mRNA)
mrna_ids_clean <- sub("\\.[0-9]+$", "", mrna_ids)
matched_symbols <- gene_map$hgnc_symbol[match(mrna_ids_clean, gene_map$ensembl_gene_id)]
mrna_labels <- ifelse(!is.na(matched_symbols) & matched_symbols != "", matched_symbols, mrna_ids)

variable_labels <- list(protein = colnames(X$protein), mRNA = mrna_labels)

stopifnot(
    length(variable_labels$protein) == ncol(X$protein),
    length(variable_labels$mRNA) == ncol(X$mRNA)
)
```

```{r}
# svg("../results/figS3b_diablo_loadings.svg", width = 8, height = 5)
plotLoadings(
    diablo.pretty, comp = 1, contrib = "max", method = "median",
    name.var = variable_labels, legend.title = "SNAEs",
    title = "Top 25 contributing variables"
)
# dev.off()
```

## Circos plot (Fig. S3C)

```{r}
# svg("../results/figS3c_diablo_circos.svg", width = 7, height = 7)
corr.pretty <- circosPlot(diablo.pretty, cutoff = 0.6, line = TRUE,
                          color.blocks = c('darkorchid', 'brown1'),
                          color.cor = c("chocolate3", "grey20"), size.labels = 1.5,
                          size.variables = 0.8, comp = 1,
                          legend.title = "SNAEs",
                          var.names = variable_labels)
# dev.off()
```

## Pairwise correlations among the 25 selected features per block (Fig. S3D)

```{r}
#| message: false
deg_group <- read_tsv("../results/tableS7_deg_group.tsv") |> 
    filter(padj < 0.05)
dep_group <- read_tsv("../results/tableS2_dep_group.tsv") |>
    filter(Threshold == "Significant")
```

```{r}
display_lookup <- c(
    setNames(variable_labels$protein, colnames(X$protein)),
    setNames(variable_labels$mRNA, colnames(X$mRNA))
)

mat <- corr.pretty[1:25, 26:50]

rn_original <- rownames(mat)
cn_original <- colnames(mat)

bold_rows <- rn_original %in% dep_group$Assay
bold_columns <- sub("\\.[0-9]+$", "", cn_original) %in% sub("\\.[0-9]+$", "", deg_group$gene)

new_rn <- unname(display_lookup[rn_original])
new_cn <- unname(display_lookup[cn_original])
new_rn[is.na(new_rn) | new_rn == ""] <- rn_original[is.na(new_rn) | new_rn == ""]
new_cn[is.na(new_cn) | new_cn == ""] <- cn_original[is.na(new_cn) | new_cn == ""]

rownames(mat) <- new_rn
colnames(mat) <- new_cn

diablo_selected_heatmap <- Heatmap(
    mat,
    name = "Pearson\ncorrelation",
    row_names_gp = gpar(fontface = ifelse(bold_rows, "bold", "plain")),
    column_names_gp = gpar(fontface = ifelse(bold_columns, "bold", "plain"))
)

# svg("../results/figS3d_diablo_selected_heatmap.svg", width = 7, height = 7)
draw(diablo_selected_heatmap)
# dev.off()
```

## Session info

```{r}
sessioninfo::session_info()
```
