---
title: "9. Correlation analysis"
format:
  html:
    df-print: paged
    embed-resources: true
    toc: true
  gfm: default
---

## Purpose

This script computes cross-omic correlations between differentially expressed proteins and genes, restricted to the SNAE comparison (Fig. 3A) and expanded to any outcome (Fig. 3B). It also correlates LGALS9 plasma NPX with HAVCR2 transcript expression (Fig. S10).

## Reproducibility

This script runs entirely from the public repository, using `olink_npx_qced_sid.tsv`, `rnaseq_transcriptomics_vst.tsv`, and the `dep_*.tsv` files.

```{r}
#| message: false
library(ComplexHeatmap)
library(circlize)
library(grid)
library(OlinkAnalyze)
library(tidyverse)
```

Import proteomics data:

```{r}
#| message: false
pdata <- read_tsv("../data/olink_npx_qced_sid.tsv")
```

Import metadata:

```{r}
#| message: false
mdata <- pdata |>
  dplyr::select(sid:death_all) |>
  distinct() |>
  arrange(sid)
```

Read differentially expressed proteins:

```{r}
#| message: false
dep_group <- read_tsv("../results/tableS2_dep_group.tsv") |>
  filter(Threshold == "Significant") |> pull(Assay)
dep_nadm <- read_tsv("../results/tableS3_dep_nadm.tsv") |>
  filter(Threshold == "Significant") |> pull(Assay)
dep_mace <- read_tsv("../results/tableS4_dep_mace.tsv") |>
  filter(Threshold == "Significant") |> pull(Assay)
dep_death_all <- read_tsv("../results/tableS5_dep_death_all.tsv") |>
  filter(Threshold == "Significant") |> pull(Assay)
dep_all <- unique(c(dep_group, dep_nadm, dep_mace, dep_death_all))
```

Read differentially expressed genes:

```{r}
#| message: false
deg_group <- read_tsv("../results/tableS7_deg_group.tsv") |>
    filter(padj < 0.05) |> 
    dplyr::select(gene, external_gene_name) |> 
    filter(!is.na(external_gene_name))
deg_nadm <- read_tsv("../results/tableS8_deg_nadm.tsv") |> 
    filter(padj < 0.05) |> 
    dplyr::select(gene, external_gene_name) |> 
    filter(!is.na(external_gene_name))
deg_mace <- read_tsv("../results/tableS9_deg_mace.tsv") |> 
    filter(padj < 0.05) |> 
    dplyr::select(gene, external_gene_name) |> 
    filter(!is.na(external_gene_name))
deg_death_all <- read_tsv("../results/tableS10_deg_death_all.tsv") |> 
    filter(padj < 0.05) |> 
    dplyr::select(gene, external_gene_name) |> 
    filter(!is.na(external_gene_name))
deg_all <- rbind(deg_group, deg_nadm, deg_mace, deg_death_all) |> distinct()
```

## Helper functions

Both panels of Fig. 3 run the identical protein x gene correlation + heatmap pipeline, differing only in which proteins/genes are included.

```{r}
compute_cross_omic_correlations <- function(protein_assays, gene_df) {
  pdata_cors <- pdata |>
    filter(Assay %in% protein_assays) |>
    dplyr::select(sid, Assay, NPX) |>
    pivot_wider(names_from = Assay, values_from = NPX) |>
    arrange(sid)

  rdata_cors <- read_tsv("../data/rnaseq_transcriptomics_vst.tsv", show_col_types = FALSE) |>
    filter(gene_ensembl_id %in% gene_df$gene) |>
    pivot_longer(cols = -gene_ensembl_id, names_to = "sid", values_to = "vst") |>
    pivot_wider(names_from = gene_ensembl_id, values_from = vst) |>
    filter(sid %in% pdata_cors$sid) |>
    arrange(sid)

  pdata_cors <- pdata_cors |> filter(sid %in% rdata_cors$sid)

  pdata_cors <- pdata_cors |> mutate(across(-1, ~ scale(.)[, 1]))
  rdata_cors <- rdata_cors |> mutate(across(-1, ~ scale(.)[, 1]))

  rdata_cors <- rdata_cors[, -1]
  pdata_cors <- pdata_cors[, -1]

  pairs <- expand.grid(RNA = names(rdata_cors), Protein = names(pdata_cors), stringsAsFactors = FALSE)

  results <- mapply(function(rna_col, prot_col) {
    x <- rdata_cors[[rna_col]]
    y <- pdata_cors[[prot_col]]
    test <- cor.test(x, y, use = "complete.obs")
    data.frame(RNA = rna_col, Protein = prot_col, Correlation = test$estimate, P.value = test$p.value)
  }, pairs$RNA, pairs$Protein, SIMPLIFY = FALSE)

  cor_results_df <- do.call(rbind, results)
  cor_results_df$adj.p.value <- p.adjust(cor_results_df$P.value, method = "fdr")
  cor_results_df <- cor_results_df |>
    mutate(signif = case_when(
      adj.p.value < 0.001 ~ "**",
      adj.p.value < 0.01  ~ "*",
      adj.p.value < 0.05  ~ ".",
      TRUE ~ ""
    )) |>
    left_join(gene_df, by = c("RNA" = "gene"))

  cor_results_df
}

build_correlation_heatmap <- function(cor_results_df) {
  mat_df <- cor_results_df |>
    dplyr::select(external_gene_name, Protein, Correlation) |>
    pivot_wider(names_from = Protein, values_from = Correlation)
  mat <- mat_df |> dplyr::select(-1) |> as.matrix()
  rownames(mat) <- mat_df$external_gene_name

  sig_mat_df <- cor_results_df |>
    mutate(significant = ifelse(adj.p.value < 0.05, 1, 0)) |>
    dplyr::select(external_gene_name, Protein, significant) |>
    pivot_wider(names_from = Protein, values_from = significant)
  sig_mat <- sig_mat_df |> dplyr::select(-1) |> as.matrix()
  rownames(sig_mat) <- mat_df$external_gene_name

  Heatmap(
    mat,
    name = "Correlation",
    column_title = "Differentially expressed proteins",
    row_title = "Differentially expressed genes",
    column_title_side = "bottom",
    row_title_side = "right",
    column_title_gp = gpar(fontsize = 12, fontface = "bold"),
    row_title_gp = gpar(fontsize = 12, fontface = "bold"),
    na_col = "grey90",
    cell_fun = function(j, i, x, y, w, h, fill) {
      if (sig_mat[i, j]) {
        grid::grid.rect(x = x, y = y, width = w, height = h,
                         gp = grid::gpar(col = "black", fill = NA, lwd = 1))
      }
    },
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    clustering_method_rows = "complete",
    clustering_method_columns = "complete",
    show_row_dend = TRUE,
    show_column_dend = TRUE,
    row_dend_width = unit(25, "mm"),
    column_dend_height = unit(25, "mm"),
    show_row_names = TRUE,
    show_column_names = TRUE,
    row_names_gp = gpar(fontsize = 9),
    column_names_gp = gpar(fontsize = 9),
    column_names_rot = 45,
    rect_gp = gpar(col = "grey85", lwd = 0.6),
    heatmap_legend_param = list(
      title = "Spearman\ncorrelation",
      at = c(-1, -0.5, 0, 0.5, 1),
      labels = c("-1", "-0.5", "0", "0.5", "1"),
      legend_height = unit(45, "mm"),
      title_gp = gpar(fontsize = 10, fontface = "bold"),
      labels_gp = gpar(fontsize = 9)
    )
  )
}
```

## Correlations restricted to SNAE-differentially-expressed features (Fig. 3A)

```{r}
cor_results_snae <- compute_cross_omic_correlations(dep_group, deg_group)
cor_results_snae |> filter(adj.p.value < 0.05) |> arrange(Correlation)
```

```{r}
cor_results_snae |>
  filter(adj.p.value < 0.05) |>
  arrange(Correlation) |>
  dplyr::count(Protein) |>
  arrange(desc(n))
```

```{r}
ht_snae <- build_correlation_heatmap(cor_results_snae)
# svg("../results/fig3A_correlation_heatmap_snae.svg", width = 10, height = 5)
draw(ht_snae, heatmap_legend_side = "right", annotation_legend_side = "right",
     padding = unit(c(4, 4, 4, 4), "mm"))
# dev.off()
```

## Correlations across features differentially expressed in any outcome (Fig. 3B)

```{r}
cor_results_all <- compute_cross_omic_correlations(dep_all, deg_all)
cor_results_all |> filter(adj.p.value < 0.05) |> arrange(Correlation)
```

```{r}
cor_results_all |>
  filter(adj.p.value < 0.05) |>
  arrange(Correlation) |>
  dplyr::count(Protein) |>
  arrange(desc(n))
```

```{r}
ht_all <- build_correlation_heatmap(cor_results_all)
# svg("../results/fig3B_correlation_heatmap_all.svg", width = 14, height = 9)
draw(ht_all, heatmap_legend_side = "right", annotation_legend_side = "right",
     padding = unit(c(4, 4, 4, 4), "mm"))
# dev.off()
```

## LGALS9 (Olink NPX) vs. HAVCR2 (TIM-3) PBMC transcript expression (Fig. S10)

```{r}
#| message: false
gal9_npx <- pdata |>
  filter(Assay == "LGALS9") |>
  dplyr::select(sid, LGALS9 = NPX)

havcr2_vst <- read_tsv("../data/rnaseq_transcriptomics_vst.tsv", show_col_types = FALSE) |>
  filter(gene_ensembl_id == "ENSG00000135077") |> # HAVCR2 (TIM-3)
  pivot_longer(-gene_ensembl_id, names_to = "sid", values_to = "HAVCR2_vst") |>
  dplyr::select(sid, HAVCR2_vst)

gal9_havcr2 <- gal9_npx |> inner_join(havcr2_vst, by = "sid")
```

```{r}
ct <- cor.test(gal9_havcr2$LGALS9, gal9_havcr2$HAVCR2_vst, method = "spearman")
sprintf("LGALS9 vs HAVCR2 (VST): n=%d, Spearman rho=%.3f, p=%.3f",
        nrow(gal9_havcr2), unname(ct$estimate), ct$p.value)
```

```{r}
figS10 <- ggplot(gal9_havcr2, aes(x = LGALS9, y = HAVCR2_vst)) +
  geom_point(alpha = 0.7, size = 2, color = "#2C3E50") +
  geom_smooth(method = "lm", se = TRUE, color = "#E74C3C", linewidth = 0.8) +
  labs(
    title = "Galectin-9 (LGALS9) vs. HAVCR2 (TIM-3) expression",
    subtitle = sprintf("Spearman \u03c1 = %.3f, p = %.2f, n = %d", ct$estimate, ct$p.value, nrow(gal9_havcr2)),
    x = "LGALS9 plasma NPX (Olink)",
    y = "HAVCR2 PBMC expression (VST)"
  ) +
  theme_minimal(base_size = 13)

figS10
# ggsave("../results/figS10_gal9_havcr2.svg", plot = figS10, width = 6, height = 5)
```

## Session info

```{r}
sessioninfo::session_info()
```
