---
title: "10. External validation (CNICS cohort, Abelman et al. 2026)"
format:
  html:
    df-print: paged
    embed-resources: true
    toc: true
  gfm: default
---

## Purpose

This script reproduces the external validation in the CNICS cohort, using LGALS9 NPX values from Abelman et al. It produces Fig. 6 and Table 3, the unadjusted and adjusted hazard ratios by quartile.

## Reproducibility

This script fetches public data from the BioStudies repository at render time (<https://doi.org/10.6019/S-BSST2240>), so an internet connection is required, but no data is stored in this repository. The sex-adjusted sensitivity model uses a private, non-publishable CNICS extract and is included for transparency on the additional analysis without redistributing that data.

```{r}
#| message: false
library(broom)
library(ggpubr)
library(ggsurvfit)
library(gtsummary)
library(patchwork)
library(survival)
library(tidyverse)
```

```{r}
#| message: false
biostudies_base <- "https://ftp.ebi.ac.uk/biostudies/fire/S-BSST/240/S-BSST2240/Files"

pdata <- read_csv(file.path(biostudies_base, "npxdat_2025_10_14.csv")) |>
    dplyr::select(ID, ends_with(c("LGALS9", "CCL13", "DAPP1", "SKAP2"))) |>
    dplyr::select(ID, starts_with("npx"))

mdata <- read_csv(file.path(biostudies_base, "other_2025_10_14.csv"))
mdata2 <- read_csv(file.path(biostudies_base, "table1_2025_10_14.csv"))
```

```{r}
all_data <- pdata |>
    left_join(mdata, by = "ID") |>
    left_join(mdata2, by = "ID") |>
    mutate(CD4_Nadir_200 = ifelse(CD4_Nadir < 200, 1, 0)) |>
    pivot_longer(cols = starts_with("npx"), names_to = "protein", values_to = "npx",
                 names_prefix = "npx") |>
    group_by(protein) |>
    mutate(quantile = factor(ntile(npx, 4))) |>
    ungroup()
```

```{r}
gal9_data <- all_data |>
    filter(protein == "LGALS9") |>
    mutate(quantile = relevel(factor(quantile), ref = "1"))
```

## Table 3: unadjusted hazard ratios (reproducible from public data)

```{r}
# Unadjusted (categorical, per-quartile HR)
gal9_cox_crude <- coxph(Surv(followup, death) ~ quantile, data = gal9_data)

# Linear-trend (one HR per quartile increase) -- for the Global P value row
gal9_trend_crude <- coxph(Surv(followup, death) ~ as.numeric(quantile), data = gal9_data)
```

```{r}
n_events <- gal9_data |> group_by(quantile) |> dplyr::count(death) |> filter(death == 1)
n_events
```

```{r}
table3_unadjusted <- bind_rows(
    tidy(gal9_cox_crude, exponentiate = TRUE, conf.int = TRUE) |>
        filter(str_detect(term, "^quantile")) |>
        mutate(quartile = str_remove(term, "quantile"), crude_estimate = estimate,
               crude_conf.low = conf.low, crude_conf.high = conf.high, .keep = "none"),
    tibble(quartile = "1", crude_estimate = 1, crude_conf.low = NA, crude_conf.high = NA)
) |>
    left_join(
        n_events |> mutate(quartile = as.character(quantile), n_events = n, .keep = "none"),
        by = "quartile"
    ) |>
    arrange(quartile)

table3_unadjusted <- table3_unadjusted |>
    bind_rows(tibble(
        quartile = "Global P value",
        crude_estimate = tidy(gal9_trend_crude)$p.value,
        crude_conf.low = NA, crude_conf.high = NA, n_events = NA
    ))

table3_unadjusted
```

## Table 3: adjusted hazard ratios (not reproducible from public data)

```{r}
gal9_data_analytic <- read_csv("../restricted_data/CNICS_sex_extract.csv") |>
    left_join(gal9_data, by = "ID")

# sex has four categories: Female, FTM, Male, MTF, but we want to analyse sex
# assigned at birth so we will reduce them to Female and Male
gal9_data_analytic$sex <- case_when(
    gal9_data_analytic$sex == "FTM" ~ "Female",
    gal9_data_analytic$sex == "MTF" ~ "Male",
    TRUE ~ gal9_data_analytic$sex
)
```

```{r}
# Adjusted (categorical, per-quartile HR)
gal9_cox_adj <- coxph(Surv(followup, death) ~ quantile + factor(vacs3) +
                          `Race/Ethnicity` + CD4_Nadir_200 + sex, data = gal9_data_analytic)

# Linear-trend, adjusted -- for the Global P value row
gal9_trend_adj <- coxph(Surv(followup, death) ~ as.numeric(quantile) + factor(vacs3) +
                            `Race/Ethnicity` + CD4_Nadir_200 + sex, data = gal9_data_analytic)
```

```{r}
table3 <- table3_unadjusted |>
    left_join(
        bind_rows(
            tidy(gal9_cox_adj, exponentiate = TRUE, conf.int = TRUE) |>
                filter(str_detect(term, "^quantile")) |>
                mutate(quartile = str_remove(term, "quantile"), adj_estimate = estimate,
                       adj_conf.low = conf.low, adj_conf.high = conf.high, .keep = "none"),
            tibble(quartile = "1", adj_estimate = 1, adj_conf.low = NA, adj_conf.high = NA)
        ),
        by = "quartile"
    ) |>
    arrange(quartile)

table3 <- table3 |>
    filter(quartile != "Global P value") |>
    bind_rows(tibble(
        quartile = "Global P value",
        crude_estimate = tidy(gal9_trend_crude)$p.value,
        crude_conf.low = NA, crude_conf.high = NA, n_events = NA,
        adj_estimate = tidy(gal9_trend_adj)$p.value[1],
        adj_conf.low = NA, adj_conf.high = NA
    ))

table3
# table3 |> write_tsv("../results/table3.tsv")
```

## Fig. 6

```{r}
spread_y <- function(y, min_sep = 0.03, y_min = 0, y_max = 1) {
    if (length(y) <= 1) return(y)
    ord <- order(y); ys <- y[ord]
    for (i in 2:length(ys)) if (ys[i] - ys[i - 1] < min_sep) ys[i] <- ys[i - 1] + min_sep
    overflow <- max(ys) - y_max
    if (overflow > 0) {
        ys <- ys - overflow
        for (i in (length(ys) - 1):1) if (ys[i + 1] - ys[i] < min_sep) ys[i] <- ys[i + 1] - min_sep
    }
    ys <- pmin(pmax(ys, y_min), y_max)
    out <- y; out[ord] <- ys; out
}

plot_cuminc_pub <- function(fit, title, xlab = "Years", ylab = "Time-to-event prob.",
                            risktable_stats = "cum.event", x_nudge = 0.6, min_sep = 0.03,
                            right_expand = 0.2, base_size = 10, p_text = NULL) {
    p <- ggcuminc(fit) +
        labs(x = xlab, y = ylab, color = "Quartile") +
        ggtitle(title) +
        theme_bw(base_size = base_size) +
        theme(plot.title = element_text(face = "bold", hjust = 0),
              axis.title = element_text(face = "bold"), legend.position = "none")
    
    lab_df <- p$data |>
        group_by(strata) |>
        slice_max(time, n = 1, with_ties = FALSE) |>
        ungroup() |>
        mutate(strata_chr = as.character(strata), qnum = str_extract(strata_chr, "\\d+"),
               label = paste0("Q", qnum), x_label = time + x_nudge)
    lab_df <- lab_df |> mutate(y_label = spread_y(estimate, min_sep = min_sep, y_min = 0, y_max = 1))
    
    p <- p +
        geom_text(data = lab_df, aes(x = x_label, y = y_label, label = label, color = strata),
                  inherit.aes = FALSE, fontface = "bold", hjust = 0, size = 3, show.legend = FALSE) +
        coord_cartesian(clip = "off") +
        scale_x_continuous(expand = expansion(mult = c(0.02, right_expand)))
    
    if (!is.null(p_text)) {
        p <- p + annotate("text", x = Inf, y = 0.02, label = p_text, hjust = 1.1, vjust = 0,
                          size = 3.2, fontface = "plain")
    }
    p <- p + add_risktable(risktable_stats = risktable_stats)
    ggsurvfit_build(p)
}
```

```{r}
fit_gal9 <- survfit2(Surv(followup, as.factor(death)) ~ quantile, data = gal9_data)

p1 <- plot_cuminc_pub(fit_gal9, "Association between LGALS9 NPX concentrations and time to event",
                      p_text = "Global P value = <0.001")

p2 <- gal9_data |>
    mutate(death = factor(death, labels = c("Survived", "Deceased"))) |>
    ggplot(aes(x = death, y = npx, fill = death)) +
    geom_boxplot(outliers = FALSE) +
    geom_jitter(width = 0.2, alpha = 0.3) +
    labs(x = NULL, y = "LGALS9 NPX", title = "LGALS9 plasma NPX by mortality status") +
    theme_bw() +
    theme(legend.position = "none") +
    stat_compare_means(comparisons = list(c("Survived", "Deceased")), method = "t.test")

w1 <- wrap_elements(full = p1)
w2 <- wrap_elements(full = p2)

fig6 <- (w1 | w2) +
    plot_layout(widths = c(1.6, 1)) +
    plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(face = "bold", size = 12), plot.tag.position = c(0.01, 0.99))

# svg("../results/fig6_CNICS_validation.svg", width = 10, height = 7)
fig6
# dev.off()
```

## Session info

```{r}
sessioninfo::session_info()
```