---
title: "11. Statistical power calculation"
format:
  html:
    df-print: paged
    embed-resources: true
    toc: true
  gfm: default
---

## Purpose

This script reproduces the study's power calculation, the sample size justification reported in the Methods, and the power curve in Fig. S12.

## Reproducibility

This script runs entirely from the public repository. No data files are needed, only the fixed sample sizes and protein counts already stated in the manuscript.

```{r}
#| message: false
library(pwr)
```

```{r}
d_effect <- 0.7
```

## Scenario 1: originally available matched population (103 pairs, N=206, 368-protein panel)

```{r}
n1 <- 103
k1 <- 368
alpha1 <- 0.05 / k1

power1 <- pwr.t.test(n = n1, d = d_effect, sig.level = alpha1,
                      type = "two.sample", alternative = "two.sided")$power

mde1 <- pwr.t.test(n = n1, power = 0.8, sig.level = alpha1,
                    type = "two.sample", alternative = "two.sided")$d

tibble::tibble(
  n_per_group = n1, N = 2 * n1, proteins_tested = k1,
  alpha_bonferroni = alpha1,
  power_at_d0.7 = power1, power_manuscript = 0.87,
  mde_at_80pct_power = mde1, mde_manuscript = 0.66
)
```

## Scenario 2: final analytic set after QC (69 pairs, 138 participants, 326 proteins)

```{r}
n2 <- 69
k2 <- 326
alpha2 <- 0.05 / k2

power2 <- pwr.t.test(n = n2, d = d_effect, sig.level = alpha2,
                      type = "two.sample", alternative = "two.sided")$power

mde2 <- pwr.t.test(n = n2, power = 0.8, sig.level = alpha2,
                    type = "two.sample", alternative = "two.sided")$d

tibble::tibble(
  n_per_group = n2, N = 2 * n2, proteins_passing_qc = k2,
  alpha_bonferroni = alpha2,
  power_at_d0.7 = power2, power_manuscript = 0.59,
  mde_at_80pct_power = mde2, mde_manuscript = 0.81
)
```

## Sanity check against the exact figures quoted in the Methods paragraph

```{r}
checks <- data.frame(
  Quantity = c("Power, Scenario 1 (368 proteins)", "MDE @80% power, Scenario 1",
               "Power, Scenario 2 (326 proteins, post-QC)", "MDE @80% power, Scenario 2"),
  R        = round(c(power1, mde1, power2, mde2), 2),
  Methods  = c(0.87, 0.66, 0.59, 0.81)
)
checks$Match <- ifelse(abs(checks$R - checks$Methods) < 0.005, "OK", "MISMATCH")
checks
```

## Fig. S12: power vs. total sample size

d=0.7, alpha=0.05/326 (post-QC threshold, the one carried into the manuscript figure), balanced groups (n1 = n2).

```{r}
n_seq <- seq(40, 220, by = 20)
curve_power <- sapply(n_seq / 2, function(n_per_group) {
  pwr.t.test(n = n_per_group, d = d_effect, sig.level = alpha2,
             type = "two.sample", alternative = "two.sided")$power
})
curve_df <- data.frame(N_total = n_seq, N_per_group = n_seq / 2,
                        Power = round(curve_power, 4))
curve_df
```

```{r}
plot_power_curve <- function() {
  par(cex.axis = 1.3, cex.lab = 1.5, mar = c(6, 5.5, 5, 1.5))
  plot(curve_df$N_total, curve_df$Power, type = "b", pch = 19, cex = 1.3, lwd = 2, col = "#1a3a5c",
       xlab = "Total sample size (N, balanced n1 = n2)", ylab = "Power",
       sub = sprintf("Standardized between-group difference d = %.1f; alpha = 0.05/%d (Bonferroni)",
                      d_effect, k2),
       ylim = c(0, 1), cex.main = 1.7, cex.sub = 1.15)
  abline(h = 0.8, lty = 2, lwd = 1.5, col = "grey50")
  points(2 * n2, power2, pch = 17, col = "#c0392b", cex = 1.9)
  text(2 * n2, power2, labels = sprintf("  final set\n  N=%d, power=%.0f%%", 2 * n2, 100 * power2),
       pos = 4, cex = 1.15, col = "#c0392b")
  legend("bottomright", legend = "80% power threshold", lty = 2, lwd = 1.5, col = "grey50",
         bty = "n", cex = 1.15)
}

plot_power_curve()

# pdf("../results/FigS12_power_calculation.pdf", width = 10.5, height = 7)
plot_power_curve()
# dev.off()
```

## Session info

```{r}
sessioninfo::session_info()
```
